A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-06-14 papers/execution-model/APLAS 2021/paper.tex
types streamlined - all implementable in OCaml

In preparation, we need to obtain a typed handle for the contract. We
assume that \lstinline!Cl! is the module containing the contract
library. The respective library function
\lstinline!Cl.make_contract_from_hash!
takes the hash of the contract along with representations of
the types of the parameter and the storage (from module
\lstinline!Ct!). It checks the validity of the hash and the types with
the blockchain and returns a typed
handle, which is indexed with OCaml types corresponding to parameter
and storage type.
\begin{lstlisting}[language=Caml,numbers=none,basicstyle=\small]
# let auction = Cl.make_contract_hash auction_hash
#     ~parameter:(Ct.Or (Ct.Unit, Ct.Unit))
#     ~storage:(Ct.Pair (Ct.Bool, Ct.Pair (Ct.Addr, Ct.Addr))));;
val auction :
  ((unit, unit) Either.t,
   bool * (Cl.Addr.t * Cl.Addr.t)) Cl.contract
\end{lstlisting}

@@

The domain-specific types come with different guarantees. Values of
type \TPUH\ and \TPUK\ as well as \TADDR\ are not necessarily valid,
as there might be no contract associated with a hash / no account
associated with a public key. In contrast, a value of type
$\TCONTRACT\ \TYPE$ is a public hash that is verified to be associated
with a contract with parameter type $\TYPE$. Operation hashes \OPH\
are only returned from blockchain operations. As the surface language
neither contains literals of type \TOPH\ nor are there casts into
that type, all values of \TOPH\ are valid.

\begin{definition}%[State of a node]
  The \emph{state of a node} is a pair
  $\NODE = [ \EXPRS, \ACCOUNTS
  % , \OPERATIONS
  ]$, where $\EXPRS$ is a
  set of programs and $\ACCOUNTS \subseteq \TPAK \times \TPUK$  is a set of
  implicit accounts.
  % , and
  % $\OPERATIONS \subseteq \ECS \times \TOP $ is a set of pairs of
  % expression contexts and operations.
\end{definition}

@@

Moreover, there are
domain-specific primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which we describe after discussing types. An
implementation provides all of these types and operations via a library API.

Types (also in Figure~\ref{fig:syntax-expressions}) comprise a
collection of standard base types as well as 
functions, pairs, lists, sums, and option types. These types are
chosen to match with built-in types of Michelson. 
In addition, there
are domain specific types of public hashes $\TPUH$ and public keys $\TPUK$ subsumed by
a type of addresses $\TADDR$. $\TCONTRACT\ \TYPE\ \TYPEU$ is the type of
a contract with parameter type $\TYPE$ and storage type $\TYPEU$. $\CODE\ \TYPE\ \TYPEU$
indicates a Michelson program with parameter type $\TYPE$ and storage
type $\TYPEU$. Tezos tokens have type $\TTEZ$. The type $\TOPH\ \TYPE\
\TYPEU$ signifies operation hashes
returned by blockchain operations. The parameters of the hash carry
the types when originating a contract. Otherwise, they are set to the
irrelevant type $\TNO$. We take the liberty of omitting irrelevant
type parameters, that is, we write $\TOPH$ for $\TOPH\ \TNO\ \TNO$.
Querying the status of an operation
returns a value of type $\TSTATUS$. Exceptions have type
$\TEXCEPTION$.

@@

Type casts are only applicable to certain pairs of
types governed by a relation $\SubType$, which could also serve as a
subtyping relation. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, and $\inferrule{}{\TCONTRACT\ \TYPE\ \TYPEU \SubType \TPUH}$.
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter and storage type.

Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

>>>>

In preparation, we need to obtain a typed handle for the contract. We
assume that \lstinline!Cl! is the module containing the contract
library. The respective library function
\lstinline!Cl.make_contract_from_hash!
takes the hash of the contract along with representations of
the types of the parameter and the storage (from module
\lstinline!Ct!). It checks the validity of the hash and the types with
the blockchain and returns a typed
handle, which is indexed with OCaml types corresponding to parameter
and storage type.
\begin{lstlisting}[language=Caml,numbers=none,basicstyle=\small]
# let auction = Cl.make_contract_from_hash auction_hash
#     ~parameter:(Ct.Or (Ct.Unit, Ct.Unit))
#     ~storage:(Ct.Pair (Ct.Bool, Ct.Pair (Ct.Addr, Ct.Addr))));;
val auction :
  ((unit, unit) Either.t,
   bool * (Cl.Addr.t * Cl.Addr.t)) Cl.contract
\end{lstlisting}

@@

The domain-specific types come with different guarantees. Values of
type \TPUH\ and \TPUK\ as well as \TADDR\ are not necessarily valid,
as there might be no contract associated with a hash / no account
associated with a public key. In contrast, a value of type
$\TCONTRACT\ \TYPE\ \TYPEU$ is a public hash that is verified to be associated
with a contract with parameter type $\TYPE$ and storage $\TYPEU$. Operation hashes \OPH\
are only returned from blockchain operations. As the surface language
neither contains literals of type \TOPH\ nor are there casts into
that type, all values of \TOPH\ are valid.

\begin{definition}%[State of a node]
  The \emph{state of a local node} is a pair
  $\NODE = [ \EXPRS, \ACCOUNTS
  % , \OPERATIONS
  ]$, where $\EXPRS$ is a
  set of programs and $\ACCOUNTS \subseteq \TPAK \times \TPUK$  is a set of
  implicit accounts.
  % , and
  % $\OPERATIONS \subseteq \ECS \times \TOP $ is a set of pairs of
  % expression contexts and operations.
\end{definition}

@@

Moreover, there are
domain-specific primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which we describe after discussing types. An
implementation provides all of these types and operations via a library module.

Types (also in Figure~\ref{fig:syntax-expressions}) comprise a
collection of standard base types as well as 
functions, pairs, lists, sums, and option types. These types are
chosen to match with built-in types of Michelson. 
In addition, there
are domain specific types of public hashes $\TPUH$ and public keys $\TPUK$ subsumed by
a type of addresses $\TADDR$. $\TCONTRACT\ \TYPE\ \TYPEU$ is the type of
a contract with parameter type $\TYPE$ and storage type $\TYPEU$. $\CODE\ \TYPE\ \TYPEU$
indicates a Michelson program with parameter type $\TYPE$ and storage
type $\TYPEU$. Tezos tokens have type $\TTEZ$. The type $\TOPH\ \TYPE\
\TYPEU$ signifies operation hashes
returned by blockchain operations. The parameters of the hash carry
the parameter and storage types when originating a contract. Otherwise, they are set to the
uninformative type $\TNO$. We take the liberty of omitting uninformative
type parameters, that is, we write $\TOPH$ for $\TOPH\ \TNO\ \TNO$.
Querying the status of an operation
returns a value of type $\TSTATUS$. Exceptions have type
$\TEXCEPTION$.

@@

Type casts are only applicable to certain pairs of
types governed by a relation $\SubType$, which could also serve as a
subtyping relation. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, $\inferrule{}{\TCONTRACT\ \TYPE\ \TYPEU \SubType
  \TCONTRACT\ \TYPE\ \TNO}$, $\inferrule{}{\TCONTRACT\ \TYPE\ \TNO
  \SubType \TCONTRACT\ \TNO\ \TNO}$, and $\inferrule{}{\TCONTRACT\
  \TNO\ \TNO \SubType \TPUH}$. Similarly for operation hashes: $\inferrule{}{\TOPH\ \TYPE\ \TYPEU \SubType
  \TOPH\ \TYPE\ \TNO}$ and $\inferrule{}{\TOPH\ \TYPE\ \TNO
  \SubType \TOPH\ \TNO\ \TNO}$. 
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter and storage type.

We express the subtyping relation via explicit cast operations because
the host language OCaml does not support subtyping. The rationale of
the subtyping among contract types is determined by the usage
pattern. If we wish to mix uses of different contracts with the same
parameter type but different storage, then we cast them to the common
type $\TCONTRACT\ \TYPE\ \TNO$. Once we have cast a contract to type
$\TCONTRACT\ \TNO\ \TNO$, we can put it in a list with any other
contract, but the only possible operation is querying its balance
(after casting further to $\TADDR$). The function
\lstinline/make_contract_from_hash/ mentioned in
Section~\ref{sec:motivation} corresponds to a cast from $\TPUH$ to
$\TCONTRACT\ \TYPE_p\ \TYPE_s$ where $\TYPE_p$ and $\TYPE_s$ are given
by the parameters \lstinline/parameter/ and \lstinline/storage/,
respectively.  The type parameters for operation
hashes just communicate the parameters to the result of the
$\CONTRACT$ operation, so their subtyping rules are analogous to the
one for the contract type.

Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified. Figure~\ref{fig:typing-blockchain-operations} gives the
typing of contract operations and queries. A standard transfer or
contract invocation results in an uninformative operation
hash. Originating a contract results in a fully typed operation hash,
which may be turned into a typed constract handle by the $\CONTRACT$
query. We can ask for the status of any operation hash, for the
storage of a contract handle (if $\TYPE_s = \TNO$, we obtain a useless
value with uninformative type $\TNO$), and for the balance of any address.
% , and $\JTypeValue\STRING\TYPE$ for serialized values as
% stored on the blockchain. The latter judgment states $\STRING$ is
% a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}
